---
title: "Simple ETL pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple ETL pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`yamltargets` lets you describe the structure of a `{targets}` pipeline in
YAML. The YAML file says what the pipeline should do: read data, transform it,
and write outputs. The `_targets.R` file stays focused on R setup code: loading
packages, setting `{targets}` options, and defining or sourcing transform
functions.

This vignette walks through the simple ETL example included with the package.
The example reads a CSV file, applies a small cleaning function, and writes the
cleaned data as a Parquet file.

## What the example contains

The example project is installed with the package under
`inst/examples/simple-etl`. Use `system.file()` to find the installed copy:

```{r}
example_dir <- system.file("examples", "simple-etl", package = "yamltargets")
list.files(example_dir, recursive = TRUE)
```

The example has three main pieces:

| File | Purpose |
|---|---|
| `data/input.csv` | Small input dataset used by the pipeline. |
| `pipeline.yml` | Declarative pipeline specification. |
| `_targets.R` | Minimal `{targets}` entry point that loads packages, defines `clean_data()`, and returns target definitions. |

## Input data

The input data are intentionally small so the pipeline structure is easy to
see:

```{r}
input_path <- file.path(example_dir, "data", "input.csv")
readr::read_csv(input_path, show_col_types = FALSE)
```

## Pipeline YAML

The basic pipeline configuration is stored in `pipeline.yml`:

```{r}
pipeline_path <- file.path(example_dir, "pipeline.yml")
writeLines(readLines(pipeline_path))
```

The configuration is organized into three sections:

| Section | Role in the pipeline |
|---|---|
| `sources` | Define input datasets and how to read them. |
| `transforms` | Define target-to-target transformations. |
| `outputs` | Define files written from target objects. |

In this example, `raw_data` is read from `data/input.csv`, `cleaned_data` is
created by calling `clean_data(raw_data)`, and the output step writes
`cleaned_data` to `results/cleaned_data.parquet`.

For output entries, `path` is an output directory. The file name is generated
from the output `name` and `format`.

## Validate the YAML

Use `validate_pipeline()` when you want to check the YAML schema without running
the pipeline or creating target objects:

```{r}
yamltargets::validate_pipeline(pipeline_path)
```

A successful validation returns `TRUE` invisibly. If validation fails,
`yamltargets` reports the YAML section and item that need attention.

## Generate target definitions

`create_pipeline_from_yaml()` parses the YAML, validates it, and returns a list
of `{targets}` target objects. Functions named in the YAML, such as
`clean_data`, must be available when the pipeline is created and run.

```{r}
clean_data <- function(df) {
  df |>
    dplyr::filter(!is.na(id)) |>
    dplyr::mutate(dplyr::across(where(is.character), tolower))
}

targets <- yamltargets::create_pipeline_from_yaml(pipeline_path)
names(targets)
```

The generated target names reflect the pipeline structure. The source target is
`raw_data`, the transform target is `cleaned_data`, and the output target is
`save_cleaned_data`.

## The `_targets.R` file

In a real project, `{targets}` reads target definitions from `_targets.R`. The
included example keeps that file short:

```{r}
writeLines(readLines(file.path(example_dir, "_targets.R")))
```

The final expression returns `create_pipeline_from_yaml("pipeline.yml")`, so the
YAML file remains the source of truth for the pipeline graph.

## Run the example pipeline

This vignette does not run `targets::tar_make()` because vignettes are built in
package-check environments where writing a target store is unnecessary. To run
the example yourself, copy it to a writable directory and run the pipeline from
there:

```{r, eval = FALSE}
work_dir <- file.path(tempdir(), "simple-etl")
unlink(work_dir, recursive = TRUE)
file.copy(example_dir, dirname(work_dir), recursive = TRUE)

old <- setwd(work_dir)
on.exit(setwd(old), add = TRUE)

targets::tar_make()
```

After the pipeline runs, the output file will be written inside the copied
example directory:

```text
results/cleaned_data.parquet
```
